---
title: "Statistical Analysis"
author: "Elena Marochkina"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = FALSE)

```

```{r libraries, set colors}
library(tidyverse)
library(readr)
library(dplyr)
library(tidyr)
library(flextable)
library(officer)
library(RColorBrewer)
library(broom)
library(tibble)
library(nortest)
library(survival)
library(survminer)
library(ggpubr)
library(lme4)   
library(lmerTest) 
library(glmnet)


# Custom theme for visualisations
theme_custom <- theme_minimal() +
  theme (
    plot.title = element_text(size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    strip.text = element_text(size = 10),
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
  )

# Palette for visualisations
color_palette <- c(brewer.pal(9, "Pastel1"), brewer.pal(8, "Pastel2"), brewer.pal(8, "Set3"))

```

```{r Descriptive Stat functions}
# Function to calculate 95% confidence interval for the mean
ci_95 <- function(x) {
  n <- sum(!is.na(x))
  if (n < 3) return("Н/П*")
  se <- sd(x, na.rm = TRUE) / sqrt(n)  # Standard Error
  mean_x <- mean(x, na.rm = TRUE)
  ci <- c(mean_x - 1.96 * se, mean_x + 1.96 * se)
  paste0(round(ci[1], 2), " - ", round(ci[2], 2))
}

# List of descriptive statistics
statistics <- list(
  `_Number of values` = ~as.character(sum(!is.na(.x))),
  `_No data` = ~as.character(sum(is.na(.x))),
  `_Mean` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(mean(.x, na.rm = TRUE) %>% round(2))),
  `_Median` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(median(.x, na.rm = TRUE) %>% round(2))),
  `_SD` = ~ifelse(sum(!is.na(.x)) < 3, "NA", as.character(sd(.x, na.rm = TRUE) %>% round(2))),
  `_Q1 - Q3` = ~ifelse(sum(!is.na(.x)) == 0, "NA", paste0(as.character(quantile(.x, 0.25, na.rm = TRUE) %>% round(2)), " - ", as.character(quantile(.x, 0.75, na.rm = TRUE) %>% round(2)))),
  `_IQR` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(IQR(.x, na.rm = TRUE) %>% round(2))),
  `_95% CI` = ~ci_95(.x),
  `_min` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(min(.x, na.rm = TRUE) %>% round(2))),
  `_max` = ~ifelse(sum(!is.na(.x)) == 0, "NA", as.character(max(.x, na.rm = TRUE) %>% round(2)))
)

```

``` {r Statictical hypothesis check}
perform_statistical_tests <- function(data, dependent_var, independent_var, variable_name) {
  result_table <- tibble(Variable = character(),
                         Performed_test = character(),
                         Result = character(),
                         Interpretation = character())
  
  # Format p-value helper function
  format_p_value <- function(p) {
    if (p < 0.0001) {
      return("< 0.0001")  # Very small p-values
    } else {
      return(formatC(p, format = "f", digits = 4))  # Standard decimal notation
    }
  }
  
  # Check the type of independent variable
  if (is.numeric(data[[independent_var]])) {
    # If independent variable is numeric
    if (length(data[[independent_var]][!is.na(data[[independent_var]])]) > 3) {
      ad_result <- nortest::ad.test(data[[independent_var]][!is.na(data[[independent_var]])])
      
      if (ad_result$p.value > 0.05) {
        # Normal distribution
        t_test_result <- t.test(data[[independent_var]] ~ as.factor(data[[dependent_var]]), data = data)
        result_table <- result_table %>%
          add_row(
            Variable = variable_name,
            Performed_test = "T-Test",
            Result = format_p_value(t_test_result$p.value),
            Interpretation = ifelse(t_test_result$p.value < 0.05, "Significant", "Not significant")
          )
      } else {
        # Not normally distributed
        if (length(unique(data[[dependent_var]])) == 2) {
          mw_test_result <- wilcox.test(data[[independent_var]] ~ as.factor(data[[dependent_var]]), data = data)
          result_table <- result_table %>%
            add_row(
              Variable = variable_name,
              Performed_test = "Mann-Whitney U Test",
              Result = format_p_value(mw_test_result$p.value),
              Interpretation = ifelse(mw_test_result$p.value < 0.05, "Significant", "Not significant")
            )
        } else {
          stop("Dependent variable must have exactly two levels for Mann-Whitney U Test.")
        }
      }
    } else {
      stop("Not enough data points for Anderson-Darling test.")
    }
    
  } else if (is.factor(data[[independent_var]]) || is.character(data[[independent_var]])) {
    # If independent variable is categorical
    data[[dependent_var]] <- as.factor(data[[dependent_var]])
    contingency_table <- table(data[[independent_var]], data[[dependent_var]])
    
    # Check for zero counts
    if (any(contingency_table == 0)) {
      # Remove rows and columns with all zeros
      contingency_table <- contingency_table[rowSums(contingency_table) > 0, colSums(contingency_table) > 0]
      
      # If the table is now empty, stop with a warning
      if (nrow(contingency_table) == 0 || ncol(contingency_table) == 0) {
        stop("Contingency table is empty after removing zero counts.")
      }
      
      # Optionally merge low-frequency categories into "Other"
      data[[independent_var]] <- as.character(data[[independent_var]])
      low_frequency <- rownames(contingency_table)[rowSums(contingency_table) <= 5]  # Adjust threshold as needed
      data[[independent_var]][data[[independent_var]] %in% low_frequency] <- "Other"
      
      # Recalculate the table
      contingency_table <- table(data[[independent_var]], data[[dependent_var]])
    }
    
    # Perform Chi-Square Test
    chi_square_result <- chisq.test(contingency_table, simulate.p.value = TRUE)
    result_table <- result_table %>%
      add_row(
        Variable = variable_name,
        Performed_test = "Chi-Squared Test",
        Result = format_p_value(chi_square_result$p.value),
        Interpretation = ifelse(chi_square_result$p.value < 0.05, "Significant", "Not significant")
      )
  } else {
    stop("Independent variable must be either numeric or categorical.")
  }
  
  if (nrow(result_table) == 0) {
    stop("No valid tests were performed. Check your data and variables.")
  }
  
  return(result_table)
}


```

``` {r Survival Analysis}
HR_assessment <- function(survival_time, event_status, predictors, data) {
  
  # Create the formula for the Cox model
  formula <- as.formula(paste("Surv(", survival_time, ", ", event_status, ") ~", paste(predictors, collapse = " + ")))
  
  # Fit the Cox model
  cox_model <- coxph(formula, data = data)
  
  # Extract hazard ratios and confidence intervals
  hr <- exp(coef(cox_model))  # Hazard ratios
  hr_ci <- exp(confint(cox_model))  # Confidence intervals
  
  # Extract p-values
  p_values <- summary(cox_model)$coefficients[, "Pr(>|z|)"]
  
  # Get the variable names corresponding to each coefficient
  variable_names <- names(coef(cox_model))
  
  # Format p-values helper function
  format_p_value <- function(p) {
    if (is.na(p)) {
      return("NA")  # Handle missing p-values
    } else if (p < 0.0001) {
      return("< 0.0001")  # Very small p-values
    } else {
      return(formatC(p, format = "f", digits = 4))  # Standard decimal notation
    }
  }
  
  # Add interpretation logic
  interpret_hr <- function(hr) {
    if (is.na(hr)) {
      return("NA")  # Handle missing HRs
    } else if (hr > 1) {
      return("Increased risk")
    } else if (hr == 1) {
      return("No change in risk")
    } else {
      return("Decreased risk")
    }
  }
  
  # Create a tibble with hazard ratios, confidence intervals, p-values, and interpretations
  hr_table <- tibble(
    Variable = variable_names,
    Hazard_Ratio = hr,
    Lower_CI = hr_ci[, 1],
    Upper_CI = hr_ci[, 2],
    p_value = sapply(p_values, format_p_value),  # Apply p-value formatting
    Interpretation = sapply(hr, interpret_hr)   # Apply HR interpretation
  )
  
  # Return the hazard ratio table
  return(hr_table)
}

```

```{r penalized COX Regression}
# Function to run Penalized Cox Regression
run_penalized_cox <- function(data, survival_time, event_status, predictors, penalty = "lasso") {
  # Define alpha based on penalty type
  alpha_value <- ifelse(penalty == "lasso", 1, 0)  # LASSO = 1, Ridge = 0
  
  # Prepare survival object and predictors matrix
  survival_outcome <- with(data, Surv(get(survival_time), get(event_status)))
  predictor_matrix <- model.matrix(~ . -1, data = data[, predictors, drop = FALSE])
  
  # Cross-validated penalized Cox model
  cv_model <- cv.glmnet(
    x = predictor_matrix,
    y = survival_outcome,
    family = "cox",
    alpha = alpha_value
  )
  
  # Get coefficients at optimal lambda
  optimal_coefs <- coef(cv_model, s = "lambda.min")
  
  # Return key results
  list(
    optimal_lambda = cv_model$lambda.min,
    coefficients = optimal_coefs,
    significant_features = rownames(optimal_coefs)[as.matrix(optimal_coefs) != 0],
    cv_model = cv_model
  )
}
```

``` {r LOS Analysis}
analyze_predictor_LOS <- function(data, predictor, LOS, patient_id) {
  
  # Helper function to format p-values
  format_p_value <- function(p) {
    if (p < 0.0001) {
      return("< 0.0001")  # Very small p-values
    } else {
      return(formatC(p, format = "f", digits = 4))  # Standard decimal notation with 4 digits
    }
  }

  # Validate Inputs
  if (!(predictor %in% colnames(data))) {
    stop(paste("Error: Predictor column", predictor, "does not exist in the dataset."))
  }
  if (!(LOS %in% colnames(data))) {
    stop(paste("Error: LOS column", LOS, "does not exist in the dataset."))
  }
  if (!(patient_id %in% colnames(data))) {
    stop(paste("Error: Patient ID column", patient_id, "does not exist in the dataset."))
  }

  # Check if LOS is numeric
  if (!is.numeric(data[[LOS]])) {
    stop("Error: LOS column must be numeric.")
  }

  # Check predictor type
  if (is.numeric(data[[predictor]])) {
    # Numerical predictor: Fit a mixed-effects linear regression model
    model <- lmer(as.formula(paste(LOS, "~", predictor, "+ (1 |", patient_id, ")")), data = data)
    model_summary <- summary(model)

    # Extract fixed-effect estimate and p-value
    fixed_effects <- tibble(
      Variable = predictor,
      Performed_test = "Mixed-Effects Linear Model",
      Statistic = model_summary$coefficients[predictor, "Estimate"],
      P_value = format_p_value(model_summary$coefficients[predictor, "Pr(>|t|)"]),
      Interpretation = ifelse(model_summary$coefficients[predictor, "Pr(>|t|)"] < 0.05, "Significant", "Not Significant")
    )

  } else if (is.factor(data[[predictor]]) || is.character(data[[predictor]])) {
    # Categorical predictor: Fit a mixed-effects linear regression model
    data[[predictor]] <- as.factor(data[[predictor]])  # Ensure it's a factor
    model <- lmer(as.formula(paste(LOS, "~", predictor, "+ (1 |", patient_id, ")")), data = data)
    model_summary <- summary(model)

    # Extract fixed-effect estimates and p-values for all levels of the categorical predictor
    fixed_effects <- as.data.frame(model_summary$coefficients) %>%
      rownames_to_column("Term") %>%
      filter(Term != "(Intercept)") %>%
      mutate(
        Variable = predictor,
        Performed_test = "Mixed-Effects Linear Model",
        Statistic = Estimate,
        P_value = sapply(`Pr(>|t|)`, format_p_value),
        Interpretation = ifelse(`Pr(>|t|)` < 0.05, "Significant", "Not Significant")
      ) %>%
      select(Variable, Term, Performed_test, Statistic, P_value, Interpretation) %>%
      as_tibble()  # Convert to tibble
  } else {
    stop("Error: Predictor must be numeric, factor, or character.")
  }

  # Return the result table as a tibble
  return(fixed_effects)
}

```


# 1. Outcomes Data

## 1.1. Read Data

```{r read files}
patients <- read.csv("../data/raw/cleaned/PATIENTS_clean.csv", stringsAsFactors = TRUE) %>%
  mutate(across(c(SUBJECT_ID, EXPIRE_FLAG), as.factor)) %>%
  filter(AGE_AT_ADMISSION > 0)

# Create a vector of SUBJECT_IDs where age > 0
subject_ids_adults <- patients$SUBJECT_ID

# Read the ADMISSIONS dataset and filter using the SUBJECT_ID vector
admissions <- read.csv("../data/raw/cleaned/ADMISSIONS_clean.csv", stringsAsFactors = TRUE) %>%
  filter(SUBJECT_ID %in% subject_ids_adults) %>% # Filter for adult patients 
  mutate(across(c(SUBJECT_ID, HADM_ID, SUBJECT_ID_COMPOSE), as.factor))

# Read the ICUSTAYS dataset and filter using the SUBJECT_ID vector
icu_stays <- read.csv("../data/raw/cleaned/ICUSTAYS_clean.csv", stringsAsFactors = TRUE) %>%
  filter(SUBJECT_ID %in% subject_ids_adults) %>%  # Filter for adult patients
  mutate(across(c(SUBJECT_ID, HADM_ID), as.factor)) 
```

## 1.2. Desriptive Statistics and Visualisation Analysis

### 1.2.1. Outcome 1: Death Status

```{r descriptive statistics for death status}
# Death status outcome
Death_status_outcome <- admissions %>%
  select(SUBJECT_ID, SUBJECT_ID_COMPOSE, SURVIVAL_FLAG)

admissions %>%
  select(SURVIVAL_FLAG) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Value) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(Variable) %>%
  mutate(`No data` = sum(is.na(Value)),
         `% by group` = (n / sum(n)) * 100) %>%
  ungroup() %>%
  select(Variable, Value, n, `% by group`, `No data`) %>%
  arrange(Variable, Value) %>%
  flextable() %>%
  merge_v("Variable") %>%
  autofit() %>%
  border_inner(border = fp_border(color = "black", width = 1)) %>%
  border_outer(border = fp_border(color = "black", width = 1))
```

```{r barplot for death status}
# Visualization for EXPIRE_FLAG in admissions dataset
ggplot(admissions, aes(x = as.factor(SURVIVAL_FLAG), fill = as.factor(SURVIVAL_FLAG))) +
  geom_bar(color = "black") +
  geom_text(
          stat = "count", aes(label = after_stat(count)),
            size = 3, 
            fontface = "bold", 
            vjust = -0.5
    ) +
  scale_fill_manual(
    values = sample(color_palette, size = length(unique(admissions$SURVIVAL_FLAG)), replace = FALSE),
    labels = c("0" = "Not Dead", "1" = "Dead")
    ) +
  labs(
    title = "Distribution of Death Status", x = "Death Status", y = "Count", fill = "Death Status"
    ) +
  theme_custom

```

-   There were 45,007 ICU stays when patient did not die (labeled "Not Dead").
-   There were 5,735 ICU stays when patient died (labeled "Dead").

This indicates that a higher proportion of patients in the dataset survived compared to those who did not. 

### 1.2.2. Outcome 2: Length of Survival

```{r descriptive statistics for length of survival}
# Create Survival_outcome file whith unique SUBJECT_ID and summarise SURVIVAL
Survival_outcome <- admissions %>%
  select(SURVIVAL, SURVIVAL_FLAG, SUBJECT_ID, SUBJECT_ID_COMPOSE)

Survival_outcome %>%
  select(SURVIVAL) %>%
  summarise(across(everything(), statistics)) %>%
  pivot_longer(cols = everything(), names_sep = "__", names_to = c("Variable", "Stat")) %>%
  rename(`Value` = value) %>%
  flextable() %>%
  merge_v("Variable") %>%
  width(j = c("Variable", "Stat", "Value"), width = 2) %>%
  autofit() %>%
  border_inner(border = fp_border(color = "black", width = 1)) %>%
  border_outer(border = fp_border(color = "black", width = 1))
```

**Outliers Check**

``` {r outliers check for length of survival}
# Calculate mean and standard deviation
mean_value <- mean(Survival_outcome$SURVIVAL, na.rm = TRUE)
sd_value <- sd(Survival_outcome$SURVIVAL, na.rm = TRUE)

# Define the 3-sigma range
lower_bound <- mean_value - 3 * sd_value
upper_bound <- mean_value + 3 * sd_value

# Identify outliers
outliers <- Survival_outcome %>%
  filter(SURVIVAL < lower_bound | SURVIVAL > upper_bound)

# Calculate the number and percentage of outliers
num_outliers <- nrow(outliers)
total_values <- nrow(Survival_outcome)
percent_outliers <- (num_outliers / total_values) * 100

# Print the number and percentage of outliers
cat("Number of outliers:", num_outliers, "\n")
cat("Percentage of outliers:", round(percent_outliers, 2), "%\n")

# Filter out the outliers
Survival_outcome <- Survival_outcome %>%
  filter(SURVIVAL >= lower_bound & SURVIVAL <= upper_bound)

# Clean the memory
rm(outliers, mean_value, sd_value, lower_bound, upper_bound, num_outliers, percent_outliers, total_values)
```

Based on the small percentage of outliers observed, it was decided to filter them out. These outliers could potentially represent patients who were in a prolonged coma or had unusually extended hospital stays.

```{r boxplot for length of survival}
# Boxplot for TIMETODEATH
ggplot(Survival_outcome, aes(y = SURVIVAL)) +
  geom_boxplot(fill = color_palette[4], color = "black") +
  labs(title = "Boxplot of Time to Death (in Days)",
       y = "Time to Death (in days)") +
  theme_custom

```

```{r density plot for length of survival}
ggplot(Survival_outcome, aes(x = SURVIVAL)) +
  geom_density(fill = color_palette[14], alpha = 0.7) +
  labs(title = "Density Plot of Time to Death (in Days)",
       x = "Time to Death (in days)",
       y = "Density") +
  theme_custom
```

The boxplot and density plot show that most patients died within 0 to 10 days, with a median around 7 days. There are still several outliers, indicating some patients survived much longer, possibly due to unique conditions such as prolonged illness or coma.

### 1.2.3. Outcome 3: Length of ICU Stay

```{r descriptive statistics for length of ICU stay}
# LOS outcome
LOS_outcome <- icu_stays %>%
  select(SUBJECT_ID, SUBJECT_ID_COMPOSE, LOS)

icu_stays %>%
  select(LOS) %>%
  summarise(across(everything(), statistics)) %>%
  pivot_longer(cols = everything(), names_sep = "__", names_to = c("Variable", "Stat")) %>%
  rename(`Value` = value) %>%
  flextable() %>%
  merge_v("Variable") %>%
  width(j = c("Variable", "Stat", "Value"), width = 2) %>%
  autofit() %>%
  border_inner(border = fp_border(color = "black", width = 1)) %>%
  border_outer(border = fp_border(color = "black", width = 1))
```

**Check for outliers**

```{r check for outliers for length of ICU stay}
# Calculate mean and standard deviation for LOS variable
mean_los <- mean(LOS_outcome$LOS, na.rm = TRUE)
sd_los <- sd(LOS_outcome$LOS, na.rm = TRUE)

# Identify outliers using the 3-sigma rule
LOS_outcome$outlier_los <- ifelse(
  LOS_outcome$LOS > (mean_los + 3 * sd_los) | LOS_outcome$LOS < (mean_los - 3 * sd_los),
  TRUE, 
  FALSE
)

# Count the number and percentage of outliers
num_outliers_los <- sum(LOS_outcome$outlier_los, na.rm = TRUE)
percent_outliers_los <- (num_outliers_los / nrow(LOS_outcome)) * 100

# Print results
cat("Number of outliers in LOS:", num_outliers_los, "\n")
cat("Percentage of outliers in LOS:", round(percent_outliers_los, 2), "%\n")

# Filter out the outliers
LOS_outcome <- LOS_outcome %>%
  select(LOS, SUBJECT_ID, SUBJECT_ID_COMPOSE, outlier_los) %>%
  filter(outlier_los == FALSE)

# Drop the outlier column after filtering
LOS_outcome$outlier_los <- NULL

# Clean the memory
rm(mean_los, sd_los, num_outliers_los, percent_outliers_los)

```

Based on the relatively small percentage of outliers (2.29%), it was decided to filter them out from the icu_stays dataset. This filtering helps ensure a more robust analysis by excluding extreme cases that could disproportionately impact the results. It's worth noting that some of these outliers could represent unique cases, such as patients in a coma or those with extended ICU stays due to complex medical conditions.

```{r violin plot for length of ICU stay}
ggplot(LOS_outcome, aes(x = "", y = LOS)) +
  geom_violin(fill = color_palette[10], color = "black") +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  labs(title = "Violin Plot of Length of ICU Stay (LOS)",
       y = "Length of ICU Stay (Days)",
       x = "") +
  theme_minimal()

```

This plot highlights that most patients had shorter ICU stays, but there are a few cases with extended durations that could represent unique medical conditions or complications.

# 2. Patients Data

## 2.1. Read and Prepare Data

```{r read file patients}
patients <- patients %>%
  mutate(
    DOB = as.POSIXct(DOB, format = "%Y-%m-%d"),
    DOD = as.POSIXct(DOD, format = "%Y-%m-%d")
  ) %>%
  filter(SUBJECT_ID %in% subject_ids_adults) # Filter for adult patients 

skimr::skim(patients %>% select(-ROW_ID, -SUBJECT_ID))
```

## 2.2. Desriptive Statistics

### 2.2.1. Descriptive Statistics for Numerical Variables (Patient data)

```{r descriptive statistics for numerical vars}
# Summarize the statistics for each numeric variable
patients %>%
  select(-ROW_ID, -SUBJECT_ID, -EXPIRE_FLAG) %>%
  select(where(is.numeric)) %>% 
  summarise(across(everything(), statistics)) %>%
  pivot_longer(cols = everything(), names_sep = "__", names_to = c("Variable", "Stat")) %>%
  rename(`Value` = value) %>%
  flextable() %>%
  merge_v("Variable") %>%
  width(j = c("Variable", "Stat", "Value"), width = 2) %>%
  autofit() %>%
  border_inner(border = fp_border(color = "black", width = 1)) %>%
  border_outer(border = fp_border(color = "black", width = 1))
```

### 2.2.2. Descriptive Statistics for Categorical Variables (Patient data)

```{r Descriptive statistics patients for categorical vars}
# Clean and summarize the categorical data for all factor variables
patients %>%
  select(-DOB, -DOD, -SUBJECT_ID, -EXPIRE_FLAG) %>%
  select(where(is.factor)) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Value) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(Variable) %>%
  mutate(`No data` = sum(is.na(Value)),
         `% by group` = (n / sum(n)) * 100) %>%
  ungroup() %>%
  select(Variable, Value, n, `% by group`, `No data`) %>%
  arrange(Variable, Value) %>%
  flextable() %>%
  merge_v("Variable") %>%
  autofit() %>%
  border_inner(border = fp_border(color = "black", width = 1)) %>%
  border_outer(border = fp_border(color = "black", width = 1))
```

## 2.3. Visualisation Analysis by Outcome

### 2.3.1. Outcome 1: Death Status by Patients Data

```{r death status by gender}
patients_death_status <- Death_status_outcome %>%
  left_join(patients, by = "SUBJECT_ID") %>%
  select(SUBJECT_ID, AGE_AT_ADMISSION, GENDER, SURVIVAL_FLAG) %>%
  distinct() %>%
  mutate(SURVIVAL_FLAG = as.factor(SURVIVAL_FLAG))  # Convert to factor

ggplot(patients_death_status, aes(x = GENDER, fill = SURVIVAL_FLAG)) +
  geom_bar(position = position_dodge2(width = 0.9), color = "black") +
  geom_text(
    stat = 'count', aes(label = after_stat(count)),
    position = position_dodge2(width = 0.9), vjust = -0.5
  ) +
  scale_fill_manual(
    values = sample(color_palette, size = length(unique(patients_death_status$SURVIVAL_FLAG)), replace = FALSE),
    labels = c("0" = "Survive", "1" = "Died")
  ) +
  labs(
    title = "Death Status by Gender",
    x = "Gender",
    y = "Count",
    fill = "Death Status"
  ) +
  theme_custom 
```

-   Males (M) have a higher count in both survival (19,569) and death (3,068) categories compared to females.
-   Females (F) have a lower count, with 14,707 surviving and 2,652 deceased.

This suggests that while the total number of male patients is higher, the proportion of deceased individuals appears similar across genders.

```{r death status by age at admission}
ggplot(patients_death_status, aes(x = AGE_AT_ADMISSION, fill = SURVIVAL_FLAG)) +
  geom_histogram(binwidth = 5, alpha = 0.6, position = "identity", color = "black") +
  scale_fill_manual(
    values = sample(color_palette, size = length(unique(patients_death_status$SURVIVAL_FLAG))), 
    labels = c("0" = "Survive", "1" = "Died")
    ) +
  labs(title = "Death Status by Age at Admission", 
       x = "Age at Admission", 
       y = "Count", 
       fill = "Death Status") +
  theme_custom
```

The histogram shows that most patients are admitted between 50–80 years, with survival dominating across all ages. However, the proportion of deaths increases with age, especially after 75, indicating that older patients are more likely to die. Age is clearly a key factor influencing mortality.

During the exploratory data analysis, a significant spike in admissions for newborns (age 0) was observed, with the vast majority of these infants being alive. After careful consideration, it was decided to exclude the data for newborns (age 0) from our analysis.

**Rationale for Removal:**
1. Different Health Dynamics in Newborns

Newborns (age 0) have distinct health characteristics and mortality patterns compared to older children and adults. Their admissions are often related to birth-related complications, immediate post-birth care, or congenital conditions. These factors may not be directly relevant to the survival analysis intended for the rest of the population.
The survival dynamics for newborns are very different from those for older children or adults, making it difficult to generalize survival factors across the age spectrum when newborns are included.

2. Survival Bias in Newborns

The vast majority of newborns survive, which introduces a potential survival bias into the analysis. Since survival outcomes for newborns are typically skewed toward "alive" (due to medical advances in neonatal care), their inclusion could dilute the significance of other age groups and make it harder to detect meaningful survival trends.
Including newborns could potentially lead to misleading conclusions about the effect of age on survival outcomes, as the survival rate for age 0 is much higher compared to other age groups, thus overshadowing the variability observed in older patients.
Distortion of Survival Model Results:

The presence of a large proportion of patients who survive (newborns) can distort the results of a survival analysis. If survival outcomes for newborns (age 0) are predominantly "alive," they can skew the overall survival curve, potentially masking or distorting the real relationship between age and survival in older populations.
By excluding newborns, we ensure that the survival analysis reflects more accurate and relevant survival patterns across the broader population, without the confounding effect of newborn health outcomes.

### 2.3.2. Outcome 2: Length of Survival by Patients Data

```{r length of survival by age and gender}
# Merge admissions with patients data to include AGE_AT_ADMISSION
patients_survival_outcome <- Survival_outcome %>%
  left_join(patients, by = "SUBJECT_ID") %>%
  select(SUBJECT_ID, AGE_AT_ADMISSION, GENDER, SURVIVAL_FLAG, SURVIVAL) %>%
  distinct() %>%
  mutate(SURVIVAL_FLAG = as.factor(SURVIVAL_FLAG))  # Convert to factor

ggplot(patients_survival_outcome, aes(x = AGE_AT_ADMISSION, y = SURVIVAL, color = GENDER)) +
  geom_point(alpha = 0.7, size = 2, shape = 16, stroke = 0.3) +  # Scatter points
  geom_smooth(method = "lm", se = TRUE, size = 1, linetype = "solid", alpha = 0.2, colour = "grey40") +  # Add linear regression line
  labs(
    title = "Length of Survival by Age",
    x = "Age at Admission",
    y = "Time to Death (Days)",
    color = "Gender"
  ) +
  scale_color_manual(
    values = sample(color_palette, size = length(unique(patients_survival_outcome$GENDER)), replace = FALSE)
  ) +
  theme_custom +
  theme(
    panel.grid.major = element_line(color = "grey90", size = 0.5),
    legend.position = "right"
  )

# Create a new age group column based on quintiles
age_breaks <- quantile(patients_survival_outcome$AGE_AT_ADMISSION, probs = seq(0, 1, 0.2), na.rm = TRUE)
age_labels <- paste0("(", round(head(age_breaks, -1)), "-", round(tail(age_breaks, -1)), "]")

patients_survival_outcome <- patients_survival_outcome %>%
  mutate(AGE_QUINTILE = cut(AGE_AT_ADMISSION, 
                            breaks = age_breaks,
                            include.lowest = TRUE,
                            labels = age_labels))

# Plot including age quintiles as facets
ggplot(patients_survival_outcome, aes(x = GENDER, y = SURVIVAL, fill = GENDER)) +
  geom_boxplot() +
  facet_wrap(~ AGE_QUINTILE, ncol = 3) +  # Facet by age quintiles
  labs(title = "Length of Survival by Gender and Age Quintile",
       x = "Gender",
       y = "Time to Death (Days)") +
  scale_fill_manual(values = sample(color_palette, size = length(unique(patients_survival_outcome$GENDER)))) +
  theme_custom

ggplot(patients_survival_outcome, aes(x = SURVIVAL, y = as.factor(AGE_QUINTILE), fill = GENDER, color = GENDER)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge2(preserve = "single")) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1.5) +
  facet_wrap(~GENDER, scales = "free_y") +
  scale_fill_manual(values = sample(color_palette, size = length(unique(patients_survival_outcome$GENDER)), replace = FALSE)) +
  scale_color_manual(values = sample(color_palette, size = length(unique(patients_survival_outcome$GENDER)), replace = FALSE)) +
  labs(
    title = "Time to Death by Age Quintile for Males and Females",
    y = "Age Quintile",
    x = "Time to Death (Days)",
    fill = "Gender",
    color = "Gender"
  ) +
  theme_custom +
  theme(
    legend.position = "right",
    strip.text = element_text(size = 12, face = "bold")
  )

rm(age_breaks, age_labels)

```

The plots allow for the visualization of how gender and age interact to influence survival times.

*Scatter Plot: Length of Survival by Age*

- No clear correlation between age at admission and time to death. Survival time is distributed across all age groups.
- Both male and female distributions appear similar, with no noticeable trend favoring one gender.

*Faceted Scatter Plot: Time to Death by Age Quintile and Gender*

- Age Quintiles: Older age groups (e.g., 80-93) show slightly higher clustering of shorter survival times.
- Within each age group, survival patterns for males and females are similar, indicating no strong gender-based differences in survival outcomes.

### 2.3.3. Outcome 3: Length of ICU Stay by Patients Data

```{r length of ICU stay by gender and age}
# Merge icu_stays with patients data to include AGE_AT_ADMISSION
patients_los_outcome <- LOS_outcome %>%
  left_join(patients, by = "SUBJECT_ID") %>%
  select(SUBJECT_ID, SUBJECT_ID_COMPOSE, LOS, AGE_AT_ADMISSION, GENDER)

ggplot(patients_los_outcome, aes(x = AGE_AT_ADMISSION, y = LOS, color = GENDER)) +
  geom_point(alpha = 0.7, size = 2, shape = 16, stroke = 0.3) +  # Scatter points
  geom_smooth(method = "lm", se = TRUE, size = 1, linetype = "solid", alpha = 0.2, colour = "grey40") +  # Add linear regression line
  labs(
    title = "Length of Stay (LOS) by by Age",
    x = "Age at Admission",
    y = "Length of Stay (Days)",
    color = "Gender"
  ) +
  scale_color_manual(
    values = sample(color_palette, size = length(unique(patients_los_outcome$GENDER)), replace = FALSE)
  ) +
  theme_custom +
  theme(
    panel.grid.major = element_line(color = "grey90", size = 0.5),
    legend.position = "right"
  )

# Create a new age group column based on quintiles
age_breaks_los <- quantile(patients_los_outcome$AGE_AT_ADMISSION, probs = seq(0, 1, 0.2), na.rm = TRUE)
age_labels_los <- paste0("(", round(head(age_breaks_los, -1)), "-", round(tail(age_breaks_los, -1)), "]")

patients_los_outcome <- patients_los_outcome %>%
  mutate(AGE_QUINTILE = cut(AGE_AT_ADMISSION, 
                             breaks = age_breaks_los,
                             include.lowest = TRUE,
                             labels = age_labels_los))

# Plot including age quintiles as facets for Length of Stay (LOS)
ggplot(patients_los_outcome, aes(x = GENDER, y = LOS, fill = GENDER)) +
  geom_boxplot() +
  facet_wrap(~ AGE_QUINTILE, ncol = 3) +
  labs(title = "Length of Stay (LOS) by Gender and Age Quintile",
       x = "Gender",
       y = "Length of Stay (Days)") +
  scale_fill_manual(values = sample(color_palette, size = length(unique(patients_los_outcome$GENDER)))) +
  theme_custom

rm(age_breaks_los, age_labels_los)
```

There is no strong trend indicating that age at admission directly correlates with the length of Stay (Days).
Gender Difference: Both genders (F and M) show similar distributions of length of Stay (Days) across all ages.

## 2.4. Exploratory analysis

### 2.4.1. Outcome 1: Death Status by Patients Data

```{r Death Status by Patients Data Exploratory analysis}
# Perform the statistical tests for both variables
result_age <- perform_statistical_tests(patients_death_status, "SURVIVAL_FLAG", "AGE_AT_ADMISSION", "Age")
result_sex <- perform_statistical_tests(patients_death_status, "SURVIVAL_FLAG", "GENDER", "Gender")

# Merge the results
merged_results <- bind_rows(result_age, result_sex)

print(merged_results)

rm(patients_death_status, result_age, result_sex, merged_results)
```

*Age:* The Mann-Whitney U Test result (< 0.0001) indicates a significant difference in death status based on age.
*Gender:* The Chi-Squared Test result (0.0005) shows a significant association between gender and death status.

Both **Age and Gender** show highly significant associations with death status (SURVIVAL_FLAG), suggesting they are both important variables in explaining survival outcomes in this dataset.

### 2.4.2. Outcome 2: Length of Survival by Patients Data

```{r Length of Survival by Patients DataExploratory analysis}
patients_survival_outcome <- patients_survival_outcome %>%
  mutate(
    SURVIVAL_FLAG = as.numeric(SURVIVAL_FLAG)
  )
  
# Generate Kaplan-Meier survival curve by GENDER
fit_gender <- survfit(Surv(SURVIVAL, SURVIVAL_FLAG) ~ GENDER, data = patients_survival_outcome)

# Plot the Kaplan-Meier curve
ggsurvplot(
  fit_gender,
  data = patients_survival_outcome,
  pval = TRUE,  # Display p-value for age groups
  conf.int = TRUE,  # Add confidence intervals
  ggtheme = theme_custom,
  surv.median.line = "hv",
  legend.title = "Age Group",
  title = "Survival Curve by Age Group",
  palette = color_palette
)

# Create age groups based on AGE_AT_ADMISSION
patients_survival_outcome <- patients_survival_outcome %>%
  mutate(
    AGE_GROUP = cut(AGE_AT_ADMISSION,
                    breaks = c(0, 18, 40, 60, 100), 
                    labels = c("0-18", "19-40", "41-60", "61+"),
                    right = FALSE)  # Include the left boundary in each group
  )

# Fit the Kaplan-Meier survival curve by age group
fit_age_group <- survfit(Surv(SURVIVAL, SURVIVAL_FLAG) ~ AGE_GROUP, data = patients_survival_outcome)

# Plot the Kaplan-Meier curve for the age groups
ggsurvplot(
  fit_age_group,
  data = patients_survival_outcome,
  pval = TRUE,  # Display p-value for age groups
  conf.int = TRUE,  # Add confidence intervals
  ggtheme = theme_custom,
  surv.median.line = "hv",
  legend.title = "Age Group",
  title = "Survival Curve by Age Group",
  palette = color_palette
)

hr_table <- HR_assessment("SURVIVAL", "SURVIVAL_FLAG", c("GENDER", "AGE_AT_ADMISSION"), patients_survival_outcome)

print(hr_table)

rm(hr_table, fit_age_group, fit_gender, patients_survival_outcome)
```

*Age at Admission:* This factor has a strong and highly significant association with mortality (p < 0.0001). Its hazard ratio (HR = 1.026) indicates that the risk of death increases with age, making it a crucial variable for predictive models.

*Gender:* Although statistically significant (p = 0.0128), the hazard ratio (HR = 0.935) suggests a small effect size. You could include gender in the model, especially if other analyses or prior knowledge suggest its importance as a confounder or interacting variable.

### 2.4.3. Outcome 3: Length of ICU Stay by Patients Data

``` {r Length of ICU Stay by Patients Data}

result_age <- analyze_predictor_LOS(patients_los_outcome, predictor = "AGE_AT_ADMISSION", LOS = "LOS", patient_id = "SUBJECT_ID")
result_sex <- analyze_predictor_LOS(patients_los_outcome, predictor = "GENDER", LOS = "LOS", patient_id = "SUBJECT_ID")

# Merge the results
merged_results <- bind_rows(result_age, result_sex)
print(merged_results)

rm(patients, merged_results, result_age, result_sex, patients_los_outcome)
```

*Age at Admission:*

- The mixed-effects linear model indicates a significant association between Age at Admission and Length of Stay in ICU (p < 0.0001).
- The positive coefficient (Statistic = 0.0053) suggests that older patients tend to have slightly longer ICU stays.

*Gender:*

- The model does not show a significant association between Gender and Length of Stay in ICU (p = 0.6624).
- This indicates that gender is unlikely to influence ICU stay duration in this dataset.

# 3. Admissions Data

## 3.1. Read and Prepare Data

```{r read file}
skimr::skim(admissions %>%
              select(-HADM_ID, -SUBJECT_ID, -SUBJECT_ID_COMPOSE) %>%
            mutate(
              SURVIVAL_FLAG = as.factor(SURVIVAL_FLAG)
            ))
```

## 3.2. Desriptive statistics

### 3.2.1. Descriptive statistics for categorical variables (Admission data)

```{r descriptive statistica admissions}
# Clean and summarize the categorical data for all factor variables
admissions %>%
  select(ADMISSION_TYPE, INSURANCE, RELIGION, MARITAL_STATUS, ETHNICITY) %>%
  select(where(is.factor)) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Value) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(Variable) %>%
  mutate(`No data` = sum(is.na(Value)),
         `% by group` = (n / sum(n)) * 100) %>%
  ungroup() %>%
  select(Variable, Value, n, `% by group`, `No data`) %>%
  arrange(Variable, Value) %>%
  flextable() %>%
  merge_v("Variable") %>%
  autofit() %>%
  border_inner(border = fp_border(color = "black", width = 1)) %>%
  border_outer(border = fp_border(color = "black", width = 1))
```

## 3.3. Visualisation Analysis by Outcome

### 3.3.1. Outcome 1: Death Status by Admissions Data

```{r death status by admissions data}
admissions_death_outcome <- admissions %>%
  select(SURVIVAL_FLAG, ADMISSION_TYPE, INSURANCE, RELIGION, MARITAL_STATUS, ETHNICITY)

# Prepare the data for plotting
admissions_death_outcome_long <- admissions_death_outcome %>%
  pivot_longer(cols = -SURVIVAL_FLAG, names_to = "Variable", values_to = "Value")

# List of categorical variables to plot
categorical_vars <- c("ADMISSION_TYPE", "INSURANCE", "RELIGION", "MARITAL_STATUS", "ETHNICITY")

# Loop through each variable and create a bar plot
for (var in categorical_vars) {
  p <- ggplot(admissions_death_outcome_long[admissions_death_outcome_long$Variable == var, ], aes(x = Value, fill = as.factor(SURVIVAL_FLAG))) +
    geom_bar(position = "dodge2", color = "black") +
    geom_text(stat = "count", aes(label = after_stat(count)), 
              position = position_dodge(width = 0.9), 
              vjust = -0.5, size = 3) +  # Adjust label position and size
    labs(title = paste("Death Status by", var),
         x = "Category",
         y = "Count",
         fill = "Death Status") +
    theme_custom +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
    scale_fill_manual(values = sample(color_palette, size = length(unique(admissions$SURVIVAL_FLAG))), labels = c("0" = "Survived", "1" = "Died"))
  
  # Print or save the plot
  print(p)
}

rm(p, categorical_vars, admissions_death_outcome_long)

```

### 3.3.2. Outcome 2: Length of Survival by Admissions Data

``` {r Outcome 2: Length of Survival by Admissions Data visualisation}
admissions_survival_outcome <- admissions %>%
  select(SUBJECT_ID_COMPOSE, SURVIVAL, SURVIVAL_FLAG, ADMISSION_TYPE, INSURANCE, RELIGION, MARITAL_STATUS, ETHNICITY) %>%
  mutate(SURVIVAL_FLAG = as.factor(SURVIVAL_FLAG))

# Function to create the plot
create_admission_plot <- function(data, x_var, color_var, color_palette, theme_custom) {
  ggplot(data, aes_string(x = x_var, y = "SURVIVAL", fill = x_var, color = color_var)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.7, fill = "white", shape = 21, size = 2, stroke = 0.5) +
    theme_custom +
    scale_fill_manual(values = color_palette) +
    scale_color_manual(values = c("black", "blue")) +
    labs(
      title = paste("Time to Death by", x_var),
      y = "Time to Death (Days)", x = x_var,
      color = "Survival Status"  # Add legend label
    ) +
    theme(
      legend.position = "right",  # Show the legend for SURVIVAL_FLAG
      strip.text = element_text(size = 12, face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
    )
}

# Define the admission variables
admission_vars <- c("ADMISSION_TYPE", "INSURANCE", "MARITAL_STATUS", "ETHNICITY", "RELIGION")

# Create a list to store the plots
plots <- lapply(admission_vars, function(var) {
  create_admission_plot(admissions_survival_outcome, var, "SURVIVAL_FLAG", color_palette, theme_custom)
})

# Name the plots for easy reference
names(plots) <- admission_vars
for (var_name in names(plots)) {
  cat("\nPlot for:", var_name, "\n")  # Print the variable name
  print(plots[[var_name]])
}

rm(plots, admission_vars, create_admission_plot, var_name, var)
``` 

### 3.3.3. Outcome 3: Length of ICU Stay by Admissions Data

``` {r Outcome 3: Length of ICU Stay by Admissions Data visualisation}
# Clean and prepare the data
admissions_los_outcome <- admissions %>%
  left_join(LOS_outcome, by = "SUBJECT_ID_COMPOSE") %>%
  select(SUBJECT_ID_COMPOSE, SURVIVAL, SURVIVAL_FLAG, ADMISSION_TYPE, INSURANCE, RELIGION, MARITAL_STATUS, ETHNICITY, LOS) %>%
  mutate(SURVIVAL_FLAG = as.factor(SURVIVAL_FLAG))

# Function to create the plot
create_LOS_plot <- function(data, x_var, color_var, color_palette, theme_custom) {
  ggplot(data, aes_string(x = x_var, y = "LOS", fill = x_var, color = color_var)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.4, fill = "white", shape = 21, size = 2, stroke = 0.5) +
    theme_custom +
    scale_fill_manual(values = color_palette) +
    scale_color_manual(values = c("black", "blue")) +
    labs(
      title = paste("Length of ICU Stay by", x_var),
      y = "Length of ICU Stay (Days)", x = x_var,
      color = "Survival Status"  # Add legend label
    ) +
    theme(
      legend.position = "right",  # Show the legend for SURVIVAL_FLAG
      strip.text = element_text(size = 12, face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels
    )
}

# Define the admission variables
admission_vars <- c("ADMISSION_TYPE", "INSURANCE", "MARITAL_STATUS", "ETHNICITY", "RELIGION")

# Create a list to store the plots
plots <- lapply(admission_vars, function(var) {
  create_LOS_plot(admissions_los_outcome, var, "SURVIVAL_FLAG", color_palette, theme_custom)
})

# Name the plots for easy reference
names(plots) <- admission_vars
for (var_name in names(plots)) {
  cat("\nPlot for:", var_name, "\n")  # Print the variable name
  print(plots[[var_name]])
}

rm(plots, admission_vars, create_LOS_plot, var_name, var)
```

## 3.4. Exploratory analysis
### 3.4.1. Outcome 1: Death Status by Admissions Data

```{r Death Status by Admissions Data }
admissions_death_outcome <- admissions_death_outcome %>%
  mutate(
    SURVIVAL_FLAG = as.numeric(as.character(SURVIVAL_FLAG))  # Ensure EXPIRE_FLAG is numeric
  )

# Perform the statistical tests for both variables
admissions_death_outcome$ADMISSION_TYPE <- droplevels(admissions_death_outcome$ADMISSION_TYPE)

result_ADMISSION_TYPE <- perform_statistical_tests(admissions_death_outcome, "SURVIVAL_FLAG", "ADMISSION_TYPE", "ADMISSION TYPE")
result_INSURANCE <- perform_statistical_tests(admissions_death_outcome, "SURVIVAL_FLAG", "INSURANCE", "INSURANCE")
result_RELIGION <- perform_statistical_tests(admissions_death_outcome, "SURVIVAL_FLAG", "RELIGION", "RELIGION")
result_MARITAL_STATUS <- perform_statistical_tests(admissions_death_outcome, "SURVIVAL_FLAG", "MARITAL_STATUS", "MARITAL STATUS")
result_ETHNICITY <- perform_statistical_tests(admissions_death_outcome, "SURVIVAL_FLAG", "ETHNICITY", "ETHNICITY")

# Merge the results
merged_results <- bind_rows(result_ADMISSION_TYPE, result_INSURANCE, result_RELIGION, result_MARITAL_STATUS, result_ETHNICITY)

print(merged_results)

rm(result_ADMISSION_TYPE, result_ETHNICITY, result_INSURANCE, result_MARITAL_STATUS, result_RELIGION, merged_results, admissions_death_outcome)

```

### 3.4.2. Outcome 2: Length of Survival by Admissions Data

```{r Length of Survival by Admissions Data exploratory analysis}
admissions_survival_outcome <- admissions_survival_outcome %>%
  mutate(
    SURVIVAL_FLAG = as.numeric(as.character(SURVIVAL_FLAG))  # Ensure EXPIRE_FLAG is numeric
  )

admissions_survival_outcome$ADMISSION_TYPE <- droplevels(admissions_survival_outcome$ADMISSION_TYPE)

# Generate Kaplan-Meier survival curve by GENDER
fit_admission_type <- survfit(Surv(SURVIVAL, SURVIVAL_FLAG) ~ ADMISSION_TYPE, data = admissions_survival_outcome)

# Plot the Kaplan-Meier curve
ggsurvplot(
  fit_admission_type,
  data = admissions_survival_outcome,
  pval = TRUE,  # Display p-value for gender
  conf.int = TRUE,  # Add confidence intervals
  ggtheme = theme_custom,
  surv.median.line = "hv",
  legend.title = "Admission type",
  title = "Survival Curve by Admission Type",
  palette =  color_palette
)

# Generate Kaplan-Meier survival curve by GENDER
fit_insurance <- survfit(Surv(SURVIVAL, SURVIVAL_FLAG) ~ INSURANCE, data = admissions_survival_outcome)

# Plot the Kaplan-Meier curve 
ggsurvplot(
  fit_insurance,
  data = admissions_survival_outcome,
  pval = TRUE,  # Display p-value for age groups
  conf.int = TRUE,  # Add confidence intervals
  ggtheme = theme_custom,
  surv.median.line = "hv",
  legend.title = "Insurance",
  title = "Survival Curve by Insurance",
  palette = color_palette
)

# Fit Kaplan-Meier model
fit_religion <- survfit(Surv(SURVIVAL, SURVIVAL_FLAG) ~ RELIGION, data = admissions_survival_outcome)

# Plot the Kaplan-Meier curve
ggsurvplot(
  fit_religion,
  data = admissions_survival_outcome,
  pval = TRUE,  # Display p-value
  conf.int = TRUE,  # Add confidence intervals
  ggtheme = theme_custom,
  surv.median.line = "hv",
  legend.title = "Religion",
  title = "Survival Curve by Religion",
  palette = color_palette
)

# Fit Kaplan-Meier model
fit_marital_status <- survfit(Surv(SURVIVAL, SURVIVAL_FLAG) ~ MARITAL_STATUS, data = admissions_survival_outcome)

# Plot the Kaplan-Meier curve
ggsurvplot(
  fit_marital_status,
  data = admissions_survival_outcome,
  pval = TRUE,  # Display p-value
  conf.int = TRUE,  # Add confidence intervals
  ggtheme = theme_custom,
  surv.median.line = "hv",
  legend.title = "Marital Status",
  title = "Survival Curve by Marital Status",
  palette = color_palette
)

# Fit Kaplan-Meier model
fit_ethnicity <- survfit(Surv(SURVIVAL, SURVIVAL_FLAG) ~ ETHNICITY, data = admissions_survival_outcome)

# Plot the Kaplan-Meier curve
ggsurvplot(
  fit_ethnicity,
  data = admissions_survival_outcome,
  pval = TRUE,  # Display p-value
  conf.int = TRUE,  # Add confidence intervals
  ggtheme = theme_custom,
  surv.median.line = "hv",
  legend.title = "Ethnicity",
  title = "Survival Curve by Ethnicity",
  palette = color_palette
)

hr_table <- HR_assessment("SURVIVAL", "SURVIVAL_FLAG", c("ADMISSION_TYPE", "INSURANCE", "MARITAL_STATUS", "RELIGION", "ETHNICITY"), admissions_survival_outcome)

print(hr_table)

rm(admissions_survival_outcome, hr_table, fit_religion, fit_admission_type, fit_ethnicity, fit_insurance, fit_marital_status)

```

### 3.4.3. Outcome 3: Length of ICU Stay by Admissions Data

``` {r Length of ICU Stay by Admission Data}
admissions_los_outcome <- admissions_los_outcome %>%
  mutate(
    SURVIVAL_FLAG = as.numeric(as.character(SURVIVAL_FLAG))
  )
  
result_ADMISSION_TYPE <- analyze_predictor_LOS(admissions_los_outcome, predictor = "ADMISSION_TYPE", LOS = "LOS", patient_id = "SUBJECT_ID_COMPOSE")
result_INSURANCE <- analyze_predictor_LOS(admissions_los_outcome, predictor = "INSURANCE", LOS = "LOS", patient_id = "SUBJECT_ID_COMPOSE")
result_RELIGION <- analyze_predictor_LOS(admissions_los_outcome, predictor = "RELIGION", LOS = "LOS", patient_id = "SUBJECT_ID_COMPOSE")
result_MARITAL_STATUS <- analyze_predictor_LOS(admissions_los_outcome, predictor = "MARITAL_STATUS", LOS = "LOS", patient_id = "SUBJECT_ID_COMPOSE")
result_ETHNICITY <- analyze_predictor_LOS(admissions_los_outcome, predictor = "ETHNICITY", LOS = "LOS", patient_id = "SUBJECT_ID_COMPOSE")

# Merge the results
merged_results <- bind_rows(result_ADMISSION_TYPE, result_INSURANCE, result_RELIGION, result_MARITAL_STATUS, result_ETHNICITY)

print(merged_results)

rm(admissions, admissions_los_outcome, merged_results, result_RELIGION, result_MARITAL_STATUS, result_ADMISSION_TYPE, result_INSURANCE, result_ETHNICITY)
```

# 4.   Diagnoses Data
## 4.1.   Read Data

``` {r read data diagnoses}
diagnoses <- read.csv("../data/raw/cleaned/DIAGNOSES_ICD_clean.csv", stringsAsFactors = TRUE) %>%
  mutate(across(where(is.character), as.factor))  %>%
  filter(SUBJECT_ID %in% subject_ids_adults) %>%
  mutate(SUBJECT_ID = as.factor(SUBJECT_ID))

skimr::skim(diagnoses %>%
              select(-SUBJECT_ID, -SUBJECT_ID_COMPOSE, -ROW_ID))
```

## 4.2.   Descriptive Statistics
### 4.2.1. Descriptive statistics for numerical variables (Diagnoses)

```{r descriptive statistics for numerical vars (Diagnoses)}
# Summarize the statistics for each numeric variable
diagnoses %>%
  select(SUBJECT_ID_COMPOSE, DIAGNOSES_NUM) %>%
  distinct() %>% # one number for 1 patient
  select(DIAGNOSES_NUM) %>%
  summarise(across(everything(), statistics)) %>%
  pivot_longer(cols = everything(), names_sep = "__", names_to = c("Variable", "Stat")) %>%
  rename(`Value` = value) %>%
  flextable() %>%
  merge_v("Variable") %>%
  width(j = c("Variable", "Stat", "Value"), width = 2) %>%
  autofit() %>%
  border_inner(border = fp_border(color = "black", width = 1)) %>%
  border_outer(border = fp_border(color = "black", width = 1))
```

### 4.2.2. Descriptive statistics for categorical variables (Diagnoses)

```{r Descriptive statistics patients for categorical vars (Diagnoses)}
# Clean and summarize the categorical data for all factor variables
diagnoses %>%
  filter(SEQ_NUM == 1) %>%
  select(where(is.factor), -SHORT_TITLE, -ICD9_CODE) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Value) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(Variable) %>%
  mutate(`No data` = sum(is.na(Value)),
         `% by group` = (n / sum(n)) * 100) %>%
  ungroup() %>%
  filter(n >= 100) %>% # Keep only rows with at least 1000 observations
  select(Variable, Value, n, `% by group`, `No data`) %>%
  arrange(Variable, desc(n), Value) %>% # Sort by Variable and descending frequency (n)
  flextable() %>%
  merge_v("Variable") %>%
  autofit() %>%
  border_inner(border = fp_border(color = "black", width = 1)) %>%
  border_outer(border = fp_border(color = "black", width = 1))

```

## 4.3. Visualisation Analysis by Outcome

### 4.3.1. Outcome 1: Death Status by Diagnoses

``` {r Death Status by Diagnoses}
# Visualize how distributes diagnoses with high-priority diagnosis
high_priority <- diagnoses %>%
  filter(SEQ_NUM == 1) %>%
  count(LONG_TITLE, sort = TRUE) %>%
  mutate(Percentage = n / sum(n) * 100,
         Cumulative_Percentage = cumsum(Percentage)) %>%
  filter(Cumulative_Percentage <= 35) # Keep diagnoses contributing to top 30%

ggplot(high_priority, aes(x = reorder(LONG_TITLE, -n), y = n)) +
  geom_bar(stat = "identity", fill = color_palette[6], color = "black") +
  geom_text(aes(label = n), hjust = 1.2, size = 3) +
  labs(title = "High-Priority Diagnoses (contributing to top 35%)",
       x = "ICD9 Code",
       y = "Frequency") +
  theme_custom +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Visualize how many diagnoses (DIAGNOSES_NUM unique number for each patients) has patients who died and who not 

diagnoses_death_outcome <- Death_status_outcome %>%
  left_join(diagnoses, by = "SUBJECT_ID_COMPOSE") %>%
  select(SUBJECT_ID_COMPOSE, SURVIVAL_FLAG, LONG_TITLE, SEQ_NUM, DIAGNOSES_NUM)

# Simplified dataset: Ensure DIAGNOSES_NUM is directly used
diagnoses_death_summary <- diagnoses_death_outcome %>%
  group_by(SURVIVAL_FLAG) %>%
  summarise(
    Mean_Diagnoses = mean(DIAGNOSES_NUM, na.rm = TRUE),
    Median_Diagnoses = median(DIAGNOSES_NUM, na.rm = TRUE),
    Count = n(),
    .groups = "drop"
  )

# Bar plot: Average number of diagnoses by death status
ggplot(diagnoses_death_summary, aes(x = factor(SURVIVAL_FLAG), y = Mean_Diagnoses, fill = factor(SURVIVAL_FLAG))) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6, color = "black") +
  geom_text(aes(label = round(Mean_Diagnoses, 1)), vjust = -0.5, size = 3) +
  labs(title = "Mean Number of Diagnoses by Death Status",
       x = "Death Status (0 = Survived, 1 = Died)",
       y = "Mean Number of Diagnoses",
       fill = "Death Status") +
  scale_fill_manual(values = sample(color_palette, size = length(unique(diagnoses_death_summary$SURVIVAL_FLAG)), replace = FALSE), labels = c("0" = "Not Dead", "1" = "Dead")) +
  theme_custom +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

rm(high_priority, diagnoses_death_summary)
```

### 4.3.2. Outcome 2: Length of Survival by Diagnoses

```{r length of survival by diagnoses}
# Merge admissions and high-priority diagnoses
top_diagnoses <- diagnoses %>%
  filter(SEQ_NUM == 1) %>%
  count(LONG_TITLE, sort = TRUE) %>%
  mutate(Percentage = n / sum(n) * 100,
         Cumulative_Percentage = cumsum(Percentage)) %>%
  filter(Cumulative_Percentage <= 35) %>%
  select(LONG_TITLE)

# Join with survival_outcome to retain survival data
diagnoses_admissions_high_priority <- Survival_outcome %>%
  left_join(diagnoses, by = "SUBJECT_ID_COMPOSE") %>% 
  filter(SEQ_NUM == 1, LONG_TITLE %in% top_diagnoses$LONG_TITLE)

# Plot: High-priority diagnosis by SURVIVAL length
ggplot(diagnoses_admissions_high_priority, aes(x = LONG_TITLE, y = SURVIVAL, fill = LONG_TITLE, color = as.factor(SURVIVAL_FLAG))) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5, fill = "white", shape = 21, size = 2, stroke = 0.5) +
  labs(title = "High-Priority Diagnoses by Survival Length",
       x = "High-Priority Diagnosis",
       y = "Survival Length (Days)") +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_custom +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
    )

# Merge admissions and diagnoses
diagnoses_survival_outcome <- Survival_outcome %>%
  left_join(diagnoses, by = "SUBJECT_ID_COMPOSE") %>%
  select(SUBJECT_ID_COMPOSE, SURVIVAL, LONG_TITLE, SEQ_NUM, DIAGNOSES_NUM)

# Plot: Average number of diagnoses per patient by survival length
diagnoses_survival_outcome %>%
  group_by(SURVIVAL) %>%
  summarise(Average_Diagnoses = mean(DIAGNOSES_NUM, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = SURVIVAL, y = Average_Diagnoses)) +
  geom_line(color = color_palette[10], size = 1.2) +  # Line for trend
  geom_point(color = color_palette[22], fill = "white", shape = 21, size = 2, stroke = 0.5) +  # Dots with outline
  geom_smooth(method = "loess", color = "darkblue", fill = "blue", alpha = 0.2, size = 1, se = TRUE) +  # LOESS smooth trend
  labs(
    title = "Average Number of Diagnoses by Survival Length",
    x = "Survival Length (Days)",
    y = "Average Number of Diagnoses"
  ) +
  scale_x_continuous(limits = c(0, max(diagnoses_survival_outcome$SURVIVAL, na.rm = TRUE)), expand = c(0, 0)) + 
  scale_y_continuous(expand = c(0, 0)) + 
  theme_custom +
  theme(
    panel.grid.major = element_line(color = "grey40", size = 0.5),  # Add gridlines
    panel.grid.minor = element_line(color = "grey95", size = 0.25)
  )

rm(top_diagnoses, diagnoses_admissions_high_priority)

```

### 4.4.3. Outcome 3: Length of ICU Stay by Diagnoses

```{r Length of ICU Stay by Diagnoses Exploratory Analysis}
# Extract top diagnoses based on frequency
top_diagnoses_los <- diagnoses %>%
  filter(SEQ_NUM == 1) %>%
  count(LONG_TITLE, sort = TRUE) %>%
  mutate(Percentage = n / sum(n) * 100,
         Cumulative_Percentage = cumsum(Percentage)) %>%
  filter(Cumulative_Percentage <= 35) %>%  # Adjust the threshold if necessary
  select(LONG_TITLE)

# Merge with LOS outcome data
diagnoses_los_outcome <- LOS_outcome %>%
  left_join(diagnoses, by = "SUBJECT_ID_COMPOSE") %>% 
  filter(SEQ_NUM == 1, LONG_TITLE %in% top_diagnoses_los$LONG_TITLE)

ggplot(diagnoses_los_outcome, aes(x = LONG_TITLE, y = LOS, fill = LONG_TITLE, color = LONG_TITLE)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5, fill = "white", shape = 21, size = 2, stroke = 0.5) +
  labs(title = "High-Priority Diagnoses by Length of ICU Stay",
       x = "High-Priority Diagnosis",
       y = "Length of ICU Stay (Days)") +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_custom +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

diagnoses_los_outcome <- LOS_outcome %>%
  left_join(diagnoses, by = "SUBJECT_ID_COMPOSE") %>%
  select(SUBJECT_ID_COMPOSE, LOS, LONG_TITLE, SEQ_NUM, DIAGNOSES_NUM)

# Plot: Average number of diagnoses per patient by LOS
diagnoses_los_outcome %>%
  group_by(LOS) %>%
  summarise(Average_Diagnoses = mean(DIAGNOSES_NUM, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = LOS, y = Average_Diagnoses)) +
  geom_line(color = color_palette[10], size = 1.2) +  # Line for trend
  geom_point(color = color_palette[22], fill = "white", shape = 21, size = 2, stroke = 0.5) +  # Dots with outline
  geom_smooth(method = "loess", color = "darkblue", fill = "blue", alpha = 0.2, size = 1, se = TRUE) +  # LOESS smooth trend
  labs(
    title = "Average Number of Diagnoses by Length of ICU Stay",
    x = "Length of ICU Stay (Days)",
    y = "Average Number of Diagnoses"
  ) +
  scale_x_continuous(limits = c(0, max(diagnoses_los_outcome$LOS, na.rm = TRUE)), expand = c(0, 0)) + 
  scale_y_continuous(expand = c(0, 0)) + 
  theme_custom +
  theme(
    panel.grid.major = element_line(color = "grey40", size = 0.5),  # Add gridlines
    panel.grid.minor = element_line(color = "grey95", size = 0.25)
  )

```

## 4.4. Exploratory analysis
### 4.4.1. Outcome 1: Death Status by Diagnoses

``` {r Death Status by Diagnoses exploratory analysis}
diagnoses_death_outcome$LONG_TITLE <- droplevels(diagnoses_death_outcome$LONG_TITLE)

result_diagnoses_num <- perform_statistical_tests(diagnoses_death_outcome, "SURVIVAL_FLAG", "DIAGNOSES_NUM", "Diagnoses number")

result_all_diagnosis <- perform_statistical_tests(diagnoses_death_outcome, "SURVIVAL_FLAG", "LONG_TITLE", "Diagnosis")

high_priority <- diagnoses_death_outcome %>%
  filter(SEQ_NUM == 1)

result_high_priority <- perform_statistical_tests(high_priority, "SURVIVAL_FLAG", "LONG_TITLE", "Key Diagnosis")

merged_results <- bind_rows(result_diagnoses_num, result_all_diagnosis, result_high_priority)

print(merged_results)

rm(result_diagnoses_num, result_all_diagnosis, high_priority, result_high_priority)
```

### 4.4.2. Outcome 2: Length of Survival by Diagnoses

```{r Length of Survival by Diagnoses Exploratory analysis}
diagnoses_survival_outcome <- Survival_outcome %>%
  left_join(diagnoses, by = "SUBJECT_ID_COMPOSE") %>%
  select(SUBJECT_ID_COMPOSE, SURVIVAL, SURVIVAL_FLAG, DIAGNOSES_NUM, LONG_TITLE, SEQ_NUM) %>%
  mutate(
    SURVIVAL_FLAG = as.numeric(as.character(SURVIVAL_FLAG)) 
  )

# Create number of diagnoses ranges based on AGE_AT_ADMISSION
diagnoses_survival_numbers <- diagnoses_survival_outcome %>%
  select(SUBJECT_ID_COMPOSE, SURVIVAL, SURVIVAL_FLAG, DIAGNOSES_NUM) %>%
  distinct() %>%
  mutate(
    DIAGNOSES_NUM = cut(DIAGNOSES_NUM,
                    breaks = c(0, 1, 3, 5, 10, 40),  # Include 0-1 as the first interval
                    labels = c("1", "1-3", "3-5", "6-10", ">10"),  # Five intervals
                    right = FALSE)  # Include the left boundary in each group
  )

# Generate Kaplan-Meier survival curve by GENDER
fit_diagnoses_num <- survfit(Surv(SURVIVAL, SURVIVAL_FLAG) ~ DIAGNOSES_NUM, data = diagnoses_survival_numbers)

# Plot the Kaplan-Meier curve
ggsurvplot(
  fit_diagnoses_num,
  data = diagnoses_survival_outcome,
  pval = TRUE,  # Display p-value for gender
  conf.int = TRUE,  # Add confidence intervals
  ggtheme = theme_custom,
  surv.median.line = "hv",
  legend.title = "Number of Diagnoses",
  title = "Survival Curve by Y",
  palette =  color_palette
)

hr_table_number_diagnoses <- HR_assessment("SURVIVAL", "SURVIVAL_FLAG", "DIAGNOSES_NUM", diagnoses_survival_numbers)

print(hr_table_number_diagnoses)

rm(diagnoses_survival_outcome, diagnoses_survival_numbers, fit_diagnoses_num, hr_table_number_diagnoses)

```

### 4.4.3. Outcome 3: Length of ICU Stay by Diagnoses

``` {r Length of ICU Stay by Admission Data}
diagnoses_LOS_numbers <- diagnoses_los_outcome %>%
  select(SUBJECT_ID_COMPOSE, LOS, DIAGNOSES_NUM) %>%
  distinct() %>%
  mutate(
    DIAGNOSES_NUM = cut(DIAGNOSES_NUM,
                    breaks = c(0, 1, 3, 5, 10, 40),  # Include 0-1 as the first interval
                    labels = c("1", "1-3", "3-5", "6-10", ">10"),  # Five intervals
                    right = FALSE)  # Include the left boundary in each group
  )

result_diagnoses_num <- analyze_predictor_LOS(diagnoses_LOS_numbers, predictor = "DIAGNOSES_NUM", LOS = "LOS", patient_id = "SUBJECT_ID_COMPOSE")

print(result_diagnoses_num)

rm(result_diagnoses_num, diagnosis_LOS_numbers, diagnoses_los_outcome)
```

